Bayesian Nonparametric System Reliability\\ using Sets of Priors

### 2015-11-17 ijar-npb-ms.tex
title suggestion and rewrite of nonpar Bayes section, reparametrisation in new section

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{url}
%\usepackage{tikz}
%\usetikzlibrary{shapes.misc,fit}

@@

%\newcommand{\comments}[1]{{\small\color{gray} #1}}

@@

@@

\begin{frontmatter}
\title{IJAR Bayesian Nonparametrics Special Issue Manuscript}

@@

The random failure time of component number $i$ of type $k$ is denoted by $T^k_i$, $i = 1, \ldots m_k$.
For its distribution, written in terms of the cdf $F^k(t) = P(T^k_i \le t)$
or in terms of the reliability function $R^k(t) = P(T^k > t) = 1 - F^k(t)$,
we consider a set of time points $t$, $t \in {\cal T} = \{t_1, \ldots, \tmax\}$.

For each time point $t$, the functioning of a single component of type $k$ at time $t$ (functioning: 1, failed: 0)
is Bernoulli distributed with a suitable functioning probability (as, $1-$failure probability) $p^k_t$, so
\begin{align*}
P\big(\indic(T^k_i > t) = 1\big) &= p^k_t\,, \\
P\big(\indic(T^k_i > t) = 0\big) &= 1 - p^k_t\,,
\end{align*}
in short, $\indic(T^k_i > t) \sim \ber(p^k_t)$, $i = 1, \ldots, m_k$.

Then, due to the assumption of independence of components of the same type, 
the number of functioning components out of the $m_k$ components of type $k$
is binomially distributed, $C^k_t = \sum_{i=1}^{m_k} \indic(T^k_i > t) \sim \bin(p^k_t, m_k)$.

The set of functioning probabilities $p^k_t$
corresponding to a choosen set of time points ($t \in {\cal T} = \{t_1, \ldots, \tmax\}$)
defines a discrete failure time distribution for components of type $k$,
$R^k(t) = p^k_{t_j}$, $t \in [t_j, t_{j+1})$.

***or say we don't know for times between time points, or lower and upper bound already here?***
***illustrate with a graph?***

%(maybe not relevant for this manuscript, but would it make sense to take Kaplan-Meier estimates for the $p^k_t$'s,
%or the NPI bounds for Kaplan-Meier?)
%

The $p^k_t$'s can be taken such that they reflect, e.g., a bathtub curve for the corresponding hazard rate.
Naturally, $p^k_{t_j} \ge p^k_{t_{j+1}}$ should hold (assuming no repair).***

@@

The posterior predictive distribution for the number of components surviving at time $t$
of the $m_k$ components in the system, based on the lifetime data and the prior information,
is then a so-called Beta-binomial distribution,
$C^k_t \mid s^k_t \sim \bebin(m_k, \alpha^k_t + s^k_t, \beta^k_t + n_k - s^k_t)$.
That is, we have
\begin{align*}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \alpha^k_t + s^k_t, m_k - l_k + \beta^k_t + n_k - s^k_t)}
                                                    {B(\alpha^k_t + s^k_t, \beta^k_t + n_k - s^k_t)} \,,
\end{align*}
where $B(\cdot, \cdot)$ is the beta function.

The posterior predictive basically averages over $p^k_t \in [0,1]$,
weighted with the posterior on $p^k_t$.

For each component type $k$, we have to choose $2 \times |{\cal T}|$
parameters to specify the prior distribution for the discrete survival function of $T^k_i$,
so in total $2 \times |{\cal T}| \times K$ parameters.

Let us drop the super- and subscript $k$ and $t$ for a while
to discuss the role and interpretation of the parameters $\alpha$ and $\beta$
for each $k$ and $t$.
The Beta distribution is usually parametrized in terms of $\alpha$ and $\beta$ as used above, 
but for our generalization to sets of priors,
it is useful to consider a different parametrization in terms of $\nz$ and $\yz$, where
\begin{align*}
\nz &= \alpha + \beta &
&\text{and} &
\yz &= \frac{\alpha}{\alpha + \beta} \,,
\end{align*}
or vice versa, $\alpha = \nz\yz$ and $\beta = \nz(1-\yz)$.
$\nz$ and $\yz$ are sometimes known as the \emph{canonical} parameters,
identified from rewriting the density in a canonical form.
(This canonical form gives a joint structure to all conjugate priors.)

From the properties of the Beta distribution,
we see that $\yz = \E[p]$ is the prior expectation for the functioning probability $p$,
and that the higher $\nz$, the more probability weight will be concentrated around that mean,
as $\V(p) = \frac{\yz (1-\yz)}{\nz + 1}$.
Furthermore, $\nz$ can be interpreted as a pseudocount or prior strength,
this will be illustrated through revisiting the posterior
based on observed lifetimes $\vec{t} = (t_1, \ldots, t_{n})$:

$p \mid s$ is Beta distributed with the updated parameters
\begin{align}
\nn &= \nz + n\,, &
\yn &= \frac{\nz}{\nz + n} \cdot \yz + \frac{n}{\nz + n} \cdot \frac{s}{n}\,,
\label{eq:nyupdate}
\end{align}
so after seeing that $s$ out of $n$ components function (at time $t$),
the posterior mean $\yn$ for $p$ is a weighted average of
the prior mean $\yz$ and $s/n$ (the fraction of function components in the data),
with the weights $\nz$ and $n$, respectively.
We see that $\nz$ plays the same role for the prior mean $\yz$
as the sample size $n$ for the observed mean $s/n$,
thus the notion of pseudocount.
Also, the higher $\nz$, the higher the weight for $\yz$
in the weighted average calculation of $\yn$,
so $\nz$ gives the strength of the prior as compared to the sample size $n$.
%The parameters $\nz$ and $\yz$ have a more intuitive interprtation,

The Beta-binomial distribution in terms of the updated parameters $\nn$ and $\yn$,
now again with indicators $t$ and $k$,
is therefore
\begin{align*}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \nn_{k,t}\yn_{k,t}, m_k - l_k + \nn_{k,t}(1-\yn_{k,t}))}
                                                    {B(\nn_{k,t}\yn_{k,t}, \nn_{k,t}(1-\yn_{k,t}))} \,.
\end{align*}

The reparametrization enables us to see that in the conjugate setting,
learning from data amounts to averaging between prior and data.
While this allows for the tractability of the model,
it also comes with a serious drawback:
When observed data are very much different from what is assumed in the prior,
this conflict is simply averaged out,
and is not reflected in the posterior or the posterior predictive.

>>>>

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{todonotes}
\usepackage{etoolbox}
\usepackage{url}
%\usepackage{tikz}
%\usetikzlibrary{shapes.misc,fit}

@@

%\newcommand{\comments}[1]{{\small\color{gray} #1}}
\newtoggle{td}
\newcommand{\td}[1]{%
  \iftoggle{td}{%
    \todo[inline]{#1}%
  }{}%
}

@@

\toggletrue{td} % show todo's
%\togglefalse{td} % hide todo's

@@

\begin{frontmatter}
\title{Bayesian Nonparametric System Reliability\\ using Sets of Priors}

@@

The random failure time of component number $i$ of type $k$ is denoted by $T^k_i$, $i = 1, \ldots m_k$.
The failure time distribution can be written in terms of the cdf $F^k(t) = P(T^k_i \le t)$,
or in terms of the reliability function $R^k(t) = P(T^k > t) = 1 - F^k(t)$,
also known as the survival function.
For a nonparametric description of $R^k(t)$,
we consider a set of time points $t$, $t \in {\cal T} = \{t_1, \ldots, \tmax\}$.

For each time point $t$, the functioning of a single component of type $k$ at time $t$ (functioning: 1, failed: 0)
is Bernoulli distributed with a suitable functioning probability (as, $1-$failure probability) $p^k_t$, so
\begin{align*}
P\big(\indic(T^k_i > t) = 1\big) &= p^k_t\,, \\
P\big(\indic(T^k_i > t) = 0\big) &= 1 - p^k_t\,,
\end{align*}
in short, $\indic(T^k_i > t) \sim \ber(p^k_t)$, $i = 1, \ldots, m_k$, $t \in {\cal T}$.

The set of functioning probabilities $\{ p^k_t, t \in {\cal T}\}$
defines a discrete failure time distribution for components of type $k$ through
\begin{align*}
R^k(t_j) &= p^k_{t_j},\ t \in {\cal T} = \{t_1, \ldots, \tmax\}\,.
\end{align*}
We can express this failure time distribution also through the probability mass function (pmf) and the discrete hazard function,
\begin{align*}
q_k(t_j) &= P(T^k = t_j) = p^k_{t_j} - p^k_{t_{j+1}}\,,\\ 
h_k(t_j) &= P(T^k = t_j \mid T^k \ge t_j) = \frac{q_k(t_j)}{R(t_j)}\,.
\end{align*}
We assume the time grid $\cal T$ is dense enough for the application at hand,
otherwise it would be possible to assume, e.g.,
$R^k(t) = p^k_{t_j}, t \in [t_j, t_{j+1})$,
or taking $p^k_{t_j}$ and $p^k_{t_{j+1}}$ as upper and lower bounds for $R^k(t)$, $t \in [t_j, t_{j+1})$.

\td{***illustrate discrete distribution with a graph of R, q, h?***}

Due to the independence assumption for components of the same type, 
the number of functioning components out of the $m_k$ components of type $k$
is binomially distributed, $C^k_t = \sum_{i=1}^{m_k} \indic(T^k_i > t) \sim \bin(p^k_t, m_k)$.
%(maybe not relevant for this manuscript, but would it make sense to take Kaplan-Meier estimates for the $p^k_t$'s,
%or the NPI bounds for Kaplan-Meier?)

The $p^k_t$'s can be taken such that they reflect, e.g., a bathtub curve for the corresponding hazard rate $h_k(t)$.
Naturally, $p^k_{t_j} \ge p^k_{t_{j+1}}$ should hold (assuming no repair).

@@

The posterior predictive distribution for the number of components surviving at time $t$
of the $m_k$ components in the system, based on the lifetime data and the prior information,
is then a so-called Beta-binomial distribution,
$C^k_t \mid s^k_t \sim \bebin(m_k, \alpha^k_t + s^k_t, \beta^k_t + n_k - s^k_t)$.
That is, we have
\begin{align*}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \alpha^k_t + s^k_t, m_k - l_k + \beta^k_t + n_k - s^k_t)}
                                                    {B(\alpha^k_t + s^k_t, \beta^k_t + n_k - s^k_t)} \,,
\end{align*}
where $B(\cdot, \cdot)$ is the Beta function.
This posterior predictive distribution is basically a Binomial distribution
where the functioning probability $p^k_t$ takes values in $[0,1]$
with the probability given by the posterior on $p^k_t$.

To wrap up, to calculate the system reliability according to ***eqnumber***,
for each component type $k$,
we need lifetime data $\vec{t}^k$,
and have to choose $2 \times |{\cal T}|$ parameters
to specify the prior distribution for the discrete survival function of $T^k_i$.
In total, values for $2 \times |{\cal T}| \times K$ parameters must be chosen.

\section{Reparametrisation of the Beta}

The parametrisation of the Beta distribution used above is common,
and allows to interpret $\alpha^k_t$ and $\beta^k_t$ as
hypothetical numbers of functioning and failed components of type $k$ at time $t$, respectively.
However, for our generalisation to sets of priors,
it is useful to consider a different parametrisation.

For discussing this reparametrisation, we will drop the super- and subscript $k$ and $t$ for a while.
Instead of $\alpha$ and $\beta$, we consider the parameters $\nz$ and $\yz$, where
\begin{align}
\nz &= \alpha + \beta &
&\text{and} &
\yz &= \frac{\alpha}{\alpha + \beta} \,,
\label{eq:reparam}
\end{align}
or vice versa, $\alpha = \nz\yz$ and $\beta = \nz(1-\yz)$.
The upper index ${}\uz$ is used to identify $\nz$ and $\yz$ as prior parameter values,
in contrast to their posterior values $\nn$ and $\yn$
obtained after observing $n$ failure times, see below.
$\nz$ and $\yz$ are sometimes called \emph{canonical} parameters,
identified from rewriting the density in a canonical form;
see, e.g., \citet[pp.~202 and 272f]{2000:bernardosmith}.
This canonical form gives a joint structure to all conjugate priors in exponential families.

From the properties of the Beta distribution,
we see that $\yz = \E[p]$ is the prior expectation for the functioning probability $p$,
and that the higher $\nz$, the more probability weight will be concentrated around $\yz$,
as $\V(p) = \frac{\yz (1-\yz)}{\nz + 1}$.
Furthermore, $\nz$ can be interpreted as a pseudocount or prior strength,
this is clear from the interpretation of $\alpha$ and $\beta$ and \eqref{eq:reparam},
and will be further illustrated below through revisiting the posterior
based on observed lifetimes $\vec{t} = (t_1, \ldots, t_{n})$.

$p \mid s$ is Beta distributed with the updated parameters
\begin{align}
\nn &= \nz + n\,, &
\yn &= \frac{\nz}{\nz + n} \cdot \yz + \frac{n}{\nz + n} \cdot \frac{s}{n}\,,
\label{eq:nyupdate}
\end{align}
so after seeing that $s$ out of $n$ components function (at time $t$),
the posterior mean $\yn$ for $p$ is a weighted average of
the prior mean $\yz$ and $s/n$ (the fraction of functioning components in the data),
with the weights $\nz$ and $n$, respectively.
We see that $\nz$ plays the same role for the prior mean $\yz$
as the sample size $n$ for the observed mean $s/n$,
thus the notion of pseudocount.
Indeed, the higher $\nz$, the higher the weight for $\yz$
in the weighted average calculation of $\yn$,
so $\nz$ gives the strength of the prior as compared to the sample size $n$.
%The parameters $\nz$ and $\yz$ have a more intuitive interprtation,

The Beta-binomial distribution in terms of the updated parameters $\nn$ and $\yn$,
now again with indicators $t$ and $k$,
is therefore
\begin{align*}
P(C^k_t = l_k \mid s^k_t) &= {m_k \choose l_k} \frac{B(l_k + \nn_{k,t}\yn_{k,t}, m_k - l_k + \nn_{k,t}(1-\yn_{k,t}))}
                                                    {B(\nn_{k,t}\yn_{k,t}, \nn_{k,t}(1-\yn_{k,t}))} \,.
\end{align*}
\td{leave this equation out???}

The parametrization in terms of prior mean and prior strength
enables us to see that in this conjugate setting,
learning from data amounts to averaging between prior and data.
While this allows for the tractability of the model,
it also comes with a serious drawback:
When observed data are very much different from what is assumed in the prior,
this conflict is simply averaged out,
and is not reflected in the posterior or the posterior predictive.
